Statistically-estimated tree composition for the northeastern United States \\ at the time of Euro-American settlement

### 2016-02-03 plos_original/paper.tex
revised and added for plos revision january 2016

\documentclass[12pt]{article}
\usepackage[]{graphicx}
\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)

%\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
%\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
%\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
%\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
%\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
%\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
%\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
%\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
%\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
%\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

%\usepackage{framed}
%\makeatletter
%\newenvironment{kframe}{%
% \def\at@end@of@kframe{}%
% \ifinner\ifhmode%
%  \def\at@end@of@kframe{\end{minipage}}%
%  \begin{minipage}{\columnwidth}%
% \fi\fi%
% \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
% \colorbox{shadecolor}{##1}\hskip-\fboxsep
%     % There is no \\@totalrightmargin, so:
%     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
% \MakeFramed {\advance\hsize-\width
%   \@totalleftmargin\z@ \linewidth\hsize
%   \@setminipage}}%
% {\par\unskip\endMakeFramed%
% \at@end@of@kframe}
%\makeatother

%\definecolor{shadecolor}{rgb}{.97, .97, .97}
%\definecolor{messagecolor}{rgb}{0, 0, 0}
%\definecolor{warningcolor}{rgb}{1, 0, 1}
%\definecolor{errorcolor}{rgb}{1, 0, 0}
%\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

@@

\makeatother

\begin{document}

@@

\lyxaddress{\textsuperscript{6}Department of Forest and Wildlife Ecology,
University of Wisconsin, Madison, Wisconsin, USA}

@@

\lyxaddress{\textsuperscript{\&}These authors contributed equally to this work.
\newpage}

\begin{abstract}
We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on a regular
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-shaped
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 0.4 under the CC BY license.

@@

\begin{figure}
\label{fig:domain}\includegraphics[scale=1.3]{fig1}

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using the full population of all the trees
in a grid cell or township.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
Data Product section) and can be used in subsequent analyses. The
mean and standard deviation of the samples for each pair of cell and
taxon represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

@@

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \citep{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with \matern
covariance \citep{Lind:etal:2011}.

@@

The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
model is this Lindgren approximation for \matern-based GPs with $\nu=1$.
To implement the Lindgren model, one modifies the $Q$ matrix defined
previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

@@

We developed an extension of the model described in previous sections
to account for data at a different aggregation than our core 8 km
grid. This extension introduces a new set of latent variables, one
per tree,  that indicate  the grid cells in which the trees are located
and can be sampled within the MCMC as additional unknown parameters.
Specifically, $c_{tj}$ is the latent ``membership'' variable for
tree $j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$
is a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
proportional to the areal overlap between the township in which the
tree is located and the $m$ grid cells, giving 
\[
c_{tj}\sim\mbox{Multinom}(1,\{\psi_{t1},\ldots,\psi_{tm}\}),
\]
independently across all trees. Because the townships overlap a limited
number of grid cells, most of the $\psi_{t1},\ldots,\psi_{tm}$ values
are zero.

@@

The model is implemented in R \citep{R:2014} with core computational
calculations coded in C++ using the \emph{Rcpp} package \citep{Edde:Fran:2011}.
We also make extensive use of sparse matrix representations and algorithms,
using the \emph{spam} package in R \citep{Furr:Sain:2010}. All code
is available on Github, including pre- and post-processing code, at
https://github.com/PalEON-Project/composition.

@@

\begin{figure}
\includegraphics[scale=0.75]{fig2}

\hspace{4mm}\includegraphics[scale=0.16]{legendRaw}\hspace{4mm}\includegraphics[scale=0.16]{legendProp}\hspace{3.5mm}\includegraphics[scale=0.16]{legendSD}

@@

\begin{figure}
\includegraphics[scale=0.8]{fig3}

@@

The authors are deeply indebted to all of the researchers over the
years who have preserved, collected, and digitized survey records,
in particular John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
We thank Madeline Ruid, Ben Seliger, Morgan Ripp and Daniel Handel
for processing of the southern Michigan data. Indiana and Illinois
data were made possible through the hard work of many Notre Dame undergraduates
in the McLachlan lab. This work was carried out by the PalEON Project
with support from the National Science Foundation MacroSystems Program
through grants EF-1065702, EF-1065656, DEB-1241874 and DEB-1241868
and from the Notre Dame Environmental Change Initiative.

@@

% \bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}
\bibliography{paperArxiv}

@@

Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
as the average of the $W$ values for the $p$th taxon in the $i$th
grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$. Let
$A$ be a diagonal matrix where $A_{ii}$ is the number of trees in
the $i$th grid cell. When there are no trees in a grid cell, $\bar{w}_{ip}=0$
and $A_{ii}=0$. For the township data, at each iteration, based on
the current values of the grid cell membership variables, $\{c_{tj}\}$,
trees are aggregated into grid cells and the calculations above can
then be carried out.

>>>>

\batchmode
\makeatletter
\def\input@path{{/accounts/gen/vis/paciorek/research/jmac/composition_paper//}}
\makeatother
\documentclass[12pt]{article}\usepackage[]{graphicx}\usepackage[]{color}
%% maxwidth is the original width if it is less than linewidth
%% otherwise use linewidth (to make sure the graphics do not exceed the margin)
\makeatletter
\def\maxwidth{ %
  \ifdim\Gin@nat@width>\linewidth
    \linewidth
  \else
    \Gin@nat@width
  \fi
}
\makeatother

\definecolor{fgcolor}{rgb}{0.345, 0.345, 0.345}
\newcommand{\hlnum}[1]{\textcolor[rgb]{0.686,0.059,0.569}{#1}}%
\newcommand{\hlstr}[1]{\textcolor[rgb]{0.192,0.494,0.8}{#1}}%
\newcommand{\hlcom}[1]{\textcolor[rgb]{0.678,0.584,0.686}{\textit{#1}}}%
\newcommand{\hlopt}[1]{\textcolor[rgb]{0,0,0}{#1}}%
\newcommand{\hlstd}[1]{\textcolor[rgb]{0.345,0.345,0.345}{#1}}%
\newcommand{\hlkwa}[1]{\textcolor[rgb]{0.161,0.373,0.58}{\textbf{#1}}}%
\newcommand{\hlkwb}[1]{\textcolor[rgb]{0.69,0.353,0.396}{#1}}%
\newcommand{\hlkwc}[1]{\textcolor[rgb]{0.333,0.667,0.333}{#1}}%
\newcommand{\hlkwd}[1]{\textcolor[rgb]{0.737,0.353,0.396}{\textbf{#1}}}%

\usepackage{framed}
\makeatletter
\newenvironment{kframe}{%
 \def\at@end@of@kframe{}%
 \ifinner\ifhmode%
  \def\at@end@of@kframe{\end{minipage}}%
  \begin{minipage}{\columnwidth}%
 \fi\fi%
 \def\FrameCommand##1{\hskip\@totalleftmargin \hskip-\fboxsep
 \colorbox{shadecolor}{##1}\hskip-\fboxsep
     % There is no \\@totalrightmargin, so:
     \hskip-\linewidth \hskip-\@totalleftmargin \hskip\columnwidth}%
 \MakeFramed {\advance\hsize-\width
   \@totalleftmargin\z@ \linewidth\hsize
   \@setminipage}}%
 {\par\unskip\endMakeFramed%
 \at@end@of@kframe}
\makeatother

\definecolor{shadecolor}{rgb}{.97, .97, .97}
\definecolor{messagecolor}{rgb}{0, 0, 0}
\definecolor{warningcolor}{rgb}{1, 0, 1}
\definecolor{errorcolor}{rgb}{1, 0, 0}
\newenvironment{knitrout}{}{} % an empty environment to be redefined in TeX

@@

\makeatother
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\begin{document}

@@

\lyxaddress{\textsuperscript{6}Department of Forestry and Wildlife Ecology,
University of Wisconsin, Madison, Wisconsin, USA}

@@

\lyxaddress{\textsuperscript{\&}These authors contributed equally to this work.
\newpage}
\begin{quote}

\end{quote}

\begin{abstract}
We present a gridded 8 km-resolution data product of the estimated
composition of tree taxa at the time of Euro-American settlement of
the northeastern United States and the statistical methodology used
to produce the product from trees recorded by land surveyors. Composition
is defined as the proportion of stems larger than approximately 20
cm diameter at breast height for 22 tree taxa, generally at the genus
level. The data come from settlement-era public survey records that
are transcribed and then aggregated spatially, giving count data.
The domain is divided into two regions, eastern (Maine to Ohio) and
midwestern (Indiana to Minnesota). Public Land Survey point data in
the midwestern region (ca. 0.8-km resolution) are aggregated to a
regular 8 km grid, while data in the eastern region, from Town Proprietor
Surveys, are aggregated at the township level in irregularly-shaped
local administrative units. The product is based on a Bayesian statistical
model fit to the count data that estimates composition on a regular
8 km grid across the entire domain. The statistical model is designed
to handle data from both the regular grid and the irregularly-sized
townships and allows us to estimate composition at locations with
no data and to smooth over noise caused by limited counts in locations
with data. Critically, the model also allows us to quantify uncertainty
in our composition estimates, making the product suitable for applications
employing data assimilation. We expect this data product to be useful
for understanding the state of vegetation in the northeastern United
States prior to large-scale Euro-American settlement. In addition
to specific regional questions, the data product can also serve as
a baseline against which to investigate how forests and ecosystems
change after intensive settlement. The data product is being made
available at the NIS data portal as version 0.4 under the CC BY license.

@@

\begin{figure}
\label{fig:domain}\includegraphics[scale=1.3]{0_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig1.pdf}

@@

We fit a Bayesian statistical model to the data, with two primary
goals:
\begin{enumerate}
\item To estimate composition on a regular grid across the entire domain,
filling gaps where no data are available, and
\item To quantify uncertainty in composition at all locations. Even in grid
cells and townships with data, we wish to quantify uncertainty because
the empirical proportions represent estimates of the true proportions
that could be calculated using the full population of all the trees
in grid cell or township.
\end{enumerate}
The result of fitting the Bayesian model via Markov chain Monte Carlo
(MCMC) is a set of representative samples from the posterior distribution
for the composition in the 23 taxonomic groupings at each of the grid
cells. These samples are the data product (described further in the
Data Product section) and can be used in subsequent analyses. The
mean and standard deviation of the samples for each pair of cell and
taxon represent our best estimate (i.e., prediction) of composition
and a Bayesian ``standard error'' quantifying the uncertainty in
the estimate.

@@

In the next section, we consider two spatial models to define the
structure of the $\alpha_{p}(s)$ processes, a standard conditional
autoregressive model \citep{Bane:etal:2004} and a Gaussian Markov
random field (MRF) approximation to a Gaussian process with Matern
covariance \citep{Lind:etal:2011}.

@@

The \citet{Lind:etal:2011} approach allows us to consider MRF approximations
to the \matern-based GP for $\nu=1$ and $\nu=2$. Our second spatial
model is the Lindgren approximation for \matern-based GPs with $\nu=1$.
To implement the Lindgren model, one modifies the $Q$ matrix defined
previously as follows. Let $a=4+\frac{1}{\rho^{2}}$. The diagonal
elements of $Q$ are $4+a^{2}$. The entries corresponding to cardinal
neighbors are $-2a$. Those for diagonal neighbors are $2$, and those
for 2nd-order cardinal neighbors are $1$. This extends the neighborhood
structure relative to the CAR model and parameterizes it as a function
of $\rho$.

@@

We developed an extension of the model described in previous sections
to account for data at a different aggregation than our core 8 km
grid. This extension introduces a new set of latent variables, one
per tree, that indicate the grid cell in which the tree is located
and can be sampled within the MCMC as additional unknown parameters.
Specifically, $c_{tj}$ is the latent ``membership'' variable for
tree $j$ in township $t$, $t=1,\ldots,T$. The prior for $c_{tj}$
is a discrete distribution that puts mass, $\psi_{ti}$, $i=1,\ldots,m$,
proportional to the areal overlap between the township in which the
tree is located and the $m$ grid cells, giving 
\[
c_{tj}\sim\mbox{Multinom}(1,\{\psi_{t1},\ldots,\psi_{tm}\}),
\]
independently across all trees. Because the townships overlap a limited
number of grid cells, most of the $\psi_{t1},\ldots,\psi_{tm}$ values
are zero.

@@

The model is implemented in R \citep{R:2014} with core computational
calculations coded in C++ using the \emph{Rcpp} package \citep{Edde:Fran:2011}.
We also make extensive use of sparse matrix representations and algorithms,
using the \emph{spam} package in R \citep{Furr:Sain:2010}. All code
is available on Github, including pre- and post-processing code, at
https://github.com/Paleon-Project/composition.

@@

\begin{figure}
\includegraphics[scale=0.75]{1_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig2.pdf}

\hspace{4mm}\includegraphics[scale=0.16]{2_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendRaw.pdf}\hspace{4mm}\includegraphics[scale=0.16]{3_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendProp.pdf}\hspace{3.5mm}\includegraphics[scale=0.16]{4_accounts_gen_vis_paciorek_research_jmac_composition_paper_legendSD.pdf}

@@

\begin{figure}
\includegraphics[scale=0.8]{5_accounts_gen_vis_paciorek_research_jmac_composition_paper_fig3.pdf}

@@

The authors are deeply indebted to all of the researchers over the
years who have preserved, collected, and digitized survey records,
in particular John Burk, Jim Dyer, Peter Marks, Robert McIntosh, Ed
Schools, Ted Sickley, Ronald Stuckey, and the Ohio Biological Survey.
We thank Madeline Ruid, Ben Seliger, Morgan Ross and Daniel Handel
for processing of the southern Michigan data. Indiana and Illinois
data were made possible through the hard work of many Notre Dame undergraduates
in the McLachlan lab. This work was carried out by the PalEON Project
with support from the National Science Foundation MacroSystems Program
through grants EF-1065702, EF-1065656, DEB-1241874 and DEB-1241868
and from the Notre Dame Environmental Change Initiative.

@@

\bibliography{/accounts/gen/vis/paciorek/bibfiles/abbrev.stat,/accounts/gen/vis/paciorek/bibfiles/abbrev.other,/accounts/gen/vis/paciorek/bibfiles/spatstat,/accounts/gen/vis/paciorek/bibfiles/statgeneral,/accounts/gen/vis/paciorek/bibfiles/paciorek,/accounts/gen/vis/paciorek/bibfiles/ecology,/accounts/gen/vis/paciorek/research/jmac/composition_paper/goring_intro}

@@

Define $\bar{w}_{ip}=\frac{1}{n_{i}}{\displaystyle \sum_{j=1}^{n_{i}}W_{ijp}}$
be the average of the $W$ values for the $p$th taxon in the $i$th
grid cell and $\bar{w}_{p}=\{\bar{w}_{ip}\},$ $i=1,\ldots,m$. Let
$A$ be a diagonal matrix where $A_{ii}$ is the number of trees in
the $i$th grid cell. When there are no trees in a grid cell, $\bar{w}_{ip}=0$
and $A_{ii}=0$. For the township data, at each iteration, based on
the current values of the grid cell membership variables, $\{c_{tj}\}$,
trees are aggregated into grid cells and the calculations above can
then be carried out.
